Reducing bias in secondary data analysis via an Explore and Confirm Analysis Workflow (ECAW): A proposal and survey of observational researchers

### 2023-04-28 manuscript/supplementary_materials.Rmd
made some cosmetic changes

# Supplementary Material B. Invitation email

## Original email

@@

## Supplementary Material C. Participant Characteristics

@@

(ref:table1c-caption) __Table C1. “What programming language or software do you use for your analyses of preexisting observational data? (you may select multiple answers)”__

@@

(ref:concernedPlotCaption) __Figure C1. Responses to survey question about number of relevant publications__

## Supplementary Material D. Figures and Tables including participants who partially complete the survey
### Participants

@@

Respondents had published a median of `r median(processed_all$n_studies, na.rm = TRUE)` (IQR `r round(quantile(processed_all$n_studies, .25, na.rm = TRUE))` to `r round(quantile(processed_all$n_studies, .75, na.rm = TRUE))`) studies using preexisting observational data (Supplementary Figure C1). They reported using the following programming languages or software packages: `r language_text_all` (Supplementary Table C1)\footnote[1]{Participants could select multiple responses to this survey question.}. `r pull(filter(concerned_all, support == "positive"), percentage)`% (`r pull(filter(concerned_all, support == "positive"), n)`/`r pull(filter(concerned_all, support == "positive"), n_sum)`) of participants reported being more concerned with research trustworthiness, bias, rigour, and reproducibility compared to what they think of as a typical research who uses preexisting observational data; `r pull(filter(concerned_all, support == "negative"), percentage)`% (`r pull(filter(concerned_all, support == "negative"), n)`/`r pull(filter(concerned_all, support == "negative"), n_sum)`) reported being less concerned (Supplementary Figure C2).

@@

```{r typicallyEcawPlotAll, warning=FALSE, fig.cap="(ref:typicallyEcawPlotAllCaption)", fig.align="center", out.width="100%", fig.width=12.8, fig.height=6, fig.path='figs/', dev=c('png', 'pdf')}
# Create typically trustworthy and reproducible plot
# Check if there is any missing and don't understand responses separately
# processed %>%
#   select(contains("typically_")) %>%
#   dplyr::filter(is.na(typically_reproducible) | is.na(typically_trustworthy) | typically_reproducible == "I don't understand the question" | typically_trustworthy == "I don't understand the question") %>%
#   nrow()

@@

(ref:typicallyEcawPlotAllCaption) __Figure 1D. Responses to the survey questions on trustworthiness and reproducibility of observational research with preexisting data and ECAWs.__ The survey defined trustworthy as “meaning that the results and conclusions of the publications are valid, reliable, rigorous, and accurate. That they merit trust”. The survey defined reproducible “in the sense that other researchers re-analysing the data with the same research question would produce similar results.” For each item, the number to the left of the data bar indicates the combined percentage for the responses depicted in any shade of brown/orange. The number in the center of the data bar (gray) indicates the percentage of neutral responses. The number to the right of the data bar indicates the combined percentage for the responses depicted in any shade of green. The bar charts in the top panel had no missing responses or selection of the option _“I don’t understand the question”_. The bar charts in the bottom panel excluded missing responses (n = `r pull(filter(ecaw_missing_all, item == "Trustworthy" & value == "Missing"), n)`; `r pull(filter(ecaw_missing_all, item == "Reproducible" & value == "Missing"), n)`) and responses of _“I don't understand the question”_ (n = `r pull(filter(ecaw_missing_all, item == "Trustworthy" & value == "I don't understand the question"), n)`; `r pull(filter(ecaw_missing_all, item == "Reproducible" & value == "I don't understand the question"), n)`).

@@

```{r methodPlotAll, warning=FALSE, fig.cap="(ref:methodPlotAllCaption)", fig.align="center", out.width="100%", fig.width=11, fig.height=4.5, fig.path='figs/', dev=c('png', 'pdf')}
# Used methods figures
method_plot_data_all <-
  processed_all %>%
  select(starts_with("method_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "scale",
  ) %>%
  separate(variable, into = c("name_prefix", "type")) %>%
  group_by(type) %>%
  count(scale) %>%
  ungroup() %>%
  replace_na(list(scale = "Missing")) %>%
  mutate(
    scale = factor(scale, levels = c(
      "Never or almost never",
      "Sometimes",
      "About half the time",
      "Most of the time",
      "Always or almost always",
      "Missing",
      "I don't understand the question"
    )),
    type = case_when(
      type == "blind" ~ "Blind the data analyst",
      type == "confirmatory" ~ "Contain confirmatory analysis",
      type == "exploratory" ~ "Contain exploratory analysis",
      type == "preregistered" ~ "Are preregistered",
      type == "script" ~ "Share analysis scripts",
    ),
    type = factor(type, levels = c(
      "Contain exploratory analysis",
      "Contain confirmatory analysis",
      "Blind the data analyst",
      "Share analysis scripts",
      "Are preregistered"
    ))
  ) %>%
  group_by(type) %>%
  tidyr::complete(scale, fill = list(n = 0)) %>%
  mutate(
    percentage = n / sum(n)
  ) %>%
  ungroup()

@@

(ref:methodPlotAllCaption) __Figure 2D. Responses to survey questions about the research practices of participants.__

@@

```{r alspacPlotAll, warning=FALSE, message=FALSE, fig.cap="(ref:alspacPlotAllCaption)", fig.align="center", out.width="100%", fig.width=14, fig.height=5, fig.path='figs/', dev=c('png', 'pdf')}
# Check if there is any missing and don't understand responses separately
alspac_missing_all <-
  processed_all %>%
  select(contains("alspac_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "item",
    values_to = "value"
  ) %>%
  dplyr::filter(is.na(value) | value == "I don't understand the question" | value == "Unsure") %>%
  mutate(
    item = str_to_title(str_remove(item, "^[^_]*_")),
    # We reorder it by making it a factor
    item = factor(item, levels = c(
      "Less_willing",
      "Opt_in",
      "Study",
      "Prefer_ecaw"
    )),
    # By making it a factor we can complete missing levels in the data later
    value = factor(value, levels = c(
      "Missing",
      "I don't understand the question",
      "Unsure"
    ))
  ) %>%
  replace_na(list(value = "Missing")) %>%
  group_by(item, value) %>%
  summarise(
    n = n()
  ) %>% 
  # Add n = 0 if no one choose a specific factor level
  tidyr::complete(value, fill = list(n = 0))

@@

(ref:alspacPlotAllCaption) __Figure 3D. Responses to survey questions about using ECAWs.__ These bar charts exclude missing values (_n_ = `r pull(filter(alspac_missing_all, item == "Less_willing" & value == "Missing"), n)`; `r pull(filter(alspac_missing_all, item == "Opt_in" & value == "Missing"), n)`; `r pull(filter(alspac_missing_all, item == "Study" & value == "Missing"), n)`; `r pull(filter(alspac_missing_all, item == "Prefer_ecaw" & value == "Missing"), n)`; respectively from top to bottom), responses of _"I don't understand the question"_ (_n_ = `r pull(filter(alspac_missing_all, item == "Less_willing" & value == "I don't understand the question"), n)`; `r pull(filter(alspac_missing_all, item == "Opt_in" & value == "I don't understand the question"), n)`; `r pull(filter(alspac_missing_all, item == "Study" & value == "I don't understand the question"), n)`; `r pull(filter(alspac_missing_all, item == "Prefer_ecaw" & value == "I don't understand the question"), n)`), and responses of _"Unsure"_ (_n_ = `r pull(filter(alspac_missing_all, item == "Less_willing" & value == "Unsure"), n)`; `r pull(filter(alspac_missing_all, item == "Opt_in" & value == "Unsure"), n)`; `r pull(filter(alspac_missing_all, item == "Study" & value == "Unsure"), n)`; `r pull(filter(alspac_missing_all, item == "Prefer_ecaw" & value == "Unsure"), n)`). Agreement with the first question may be slightly inflated due to the format of the questions in this block. Respondents with a highly positive inclination towards ECAWs would be expected to disagree with the first question, but agree with the next three questions. Four respondents agreed with all four statements, suggesting they may have glazed over the word “less” in the first question \protect\footnotemark[2]. Interpreting responses to the second and third question come with a degree of ambiguity as the survey did not specify what was meant by the term “study” \protect\footnotemark[3].

>>>>

# Supplementary Material B. Invitation email
## Original email

@@

# Supplementary Material C. Participant Characteristics

@@

```{=latex}
{\smaller[1]
\begin{singlespace}
```
(ref:table1c-caption) __Table C1. “What programming language or software do you use for your analyses of preexisting observational data? (you may select multiple answers)”__
```{=latex}
\end{singlespace}
}
```

@@

```{=latex}
{\smaller[1]
\begin{singlespace}
```
(ref:concernedPlotCaption) __Figure C1. Responses to survey question about number of relevant publications__
```{=latex}
\end{singlespace}
}
```

# Supplementary Material D. Figures and Tables including participants who partially complete the survey
## Participants

@@

Respondents had published a median of `r median(processed_all$n_studies, na.rm = TRUE)` (IQR `r round(quantile(processed_all$n_studies, .25, na.rm = TRUE))` to `r round(quantile(processed_all$n_studies, .75, na.rm = TRUE))`) studies using preexisting observational data. They reported using the following programming languages or software packages: `r language_text_all` \footnote[1]{Participants could select multiple responses to this survey question.}. `r pull(filter(concerned_all, support == "positive"), percentage)`% (`r pull(filter(concerned_all, support == "positive"), n)`/`r pull(filter(concerned_all, support == "positive"), n_sum)`) of participants reported being more concerned with research trustworthiness, bias, rigour, and reproducibility compared to what they think of as a typical research who uses preexisting observational data; `r pull(filter(concerned_all, support == "negative"), percentage)`% (`r pull(filter(concerned_all, support == "negative"), n)`/`r pull(filter(concerned_all, support == "negative"), n_sum)`) reported being less concerned.

@@

```{r typicallyEcawPlotAll, warning=FALSE, fig.cap="(ref:typicallyEcawPlotAllCaption)", fig.align="center", fig.pos="H", out.width="100%", fig.width=12.8, fig.height=6, fig.path='figs/', dev=c('png', 'pdf')}
# Create typically trustworthy and reproducible plot
# Check if there is any missing and don't understand responses separately
# processed %>%
#   select(contains("typically_")) %>%
#   dplyr::filter(is.na(typically_reproducible) | is.na(typically_trustworthy) | typically_reproducible == "I don't understand the question" | typically_trustworthy == "I don't understand the question") %>%
#   nrow()

@@

```{=latex}
{\smaller[1]
\begin{singlespace}
```
(ref:typicallyEcawPlotAllCaption) __Figure 1D. Responses to the survey questions on trustworthiness and reproducibility of observational research with preexisting data and ECAWs.__ The survey defined trustworthy as “meaning that the results and conclusions of the publications are valid, reliable, rigorous, and accurate. That they merit trust”. The survey defined reproducible “in the sense that other researchers re-analysing the data with the same research question would produce similar results.” For each item, the number to the left of the data bar indicates the combined percentage for the responses depicted in any shade of brown/orange. The number in the center of the data bar (gray) indicates the percentage of neutral responses. The number to the right of the data bar indicates the combined percentage for the responses depicted in any shade of green. The bar charts in the top panel had no missing responses or selection of the option _“I don’t understand the question”_. The bar charts in the bottom panel excluded missing responses (n = `r pull(filter(ecaw_missing_all, item == "Trustworthy" & value == "Missing"), n)`; `r pull(filter(ecaw_missing_all, item == "Reproducible" & value == "Missing"), n)`) and responses of _“I don't understand the question”_ (n = `r pull(filter(ecaw_missing_all, item == "Trustworthy" & value == "I don't understand the question"), n)`; `r pull(filter(ecaw_missing_all, item == "Reproducible" & value == "I don't understand the question"), n)`).
```{=latex}
\end{singlespace}
}
```

@@

```{r methodPlotAll, warning=FALSE, fig.cap="(ref:methodPlotAllCaption)", fig.align="center", fig.pos="H", out.width="100%", fig.width=11, fig.height=4.5, fig.path='figs/', dev=c('png', 'pdf')}
# Used methods figures
method_plot_data_all <-
  processed_all %>%
  select(starts_with("method_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "scale",
  ) %>%
  separate(variable, into = c("name_prefix", "type")) %>%
  group_by(type) %>%
  count(scale) %>%
  ungroup() %>%
  replace_na(list(scale = "Missing")) %>%
  mutate(
    scale = factor(scale, levels = c(
      "Never or almost never",
      "Sometimes",
      "About half the time",
      "Most of the time",
      "Always or almost always",
      "Missing",
      "I don't understand the question"
    )),
    type = case_when(
      type == "blind" ~ "Blind the data analyst",
      type == "confirmatory" ~ "Contain confirmatory analysis",
      type == "exploratory" ~ "Contain exploratory analysis",
      type == "preregistered" ~ "Are preregistered",
      type == "script" ~ "Share analysis scripts",
    ),
    type = factor(type, levels = c(
      "Contain exploratory analysis",
      "Contain confirmatory analysis",
      "Blind the data analyst",
      "Share analysis scripts",
      "Are preregistered"
    ))
  ) %>%
  group_by(type) %>%
  tidyr::complete(scale, fill = list(n = 0)) %>%
  mutate(
    percentage = n / sum(n)
  ) %>%
  ungroup()

@@

```{=latex}
{\smaller[1]
\begin{singlespace}
```
(ref:methodPlotAllCaption) __Figure 2D. Responses to survey questions about the research practices of participants.__
```{=latex}
\end{singlespace}
}
```

@@

```{r alspacPlotAll, warning=FALSE, message=FALSE, fig.cap="(ref:alspacPlotAllCaption)", fig.align="center", fig.pos="H", out.width="100%", fig.width=14, fig.height=5, fig.path='figs/', dev=c('png', 'pdf')}
# Check if there is any missing and don't understand responses separately
alspac_missing_all <-
  processed_all %>%
  select(contains("alspac_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "item",
    values_to = "value"
  ) %>%
  dplyr::filter(is.na(value) | value == "I don't understand the question" | value == "Unsure") %>%
  mutate(
    item = str_to_title(str_remove(item, "^[^_]*_")),
    # We reorder it by making it a factor
    item = factor(item, levels = c(
      "Less_willing",
      "Opt_in",
      "Study",
      "Prefer_ecaw"
    )),
    # By making it a factor we can complete missing levels in the data later
    value = factor(value, levels = c(
      "Missing",
      "I don't understand the question",
      "Unsure"
    ))
  ) %>%
  replace_na(list(value = "Missing")) %>%
  group_by(item, value) %>%
  summarise(
    n = n()
  ) %>% 
  # Add n = 0 if no one choose a specific factor level
  tidyr::complete(value, fill = list(n = 0))

@@

```{=latex}
{\smaller[1]
\begin{singlespace}
```
(ref:alspacPlotAllCaption) __Figure 3D. Responses to survey questions about using ECAWs.__ These bar charts exclude missing values (_n_ = `r pull(filter(alspac_missing_all, item == "Less_willing" & value == "Missing"), n)`; `r pull(filter(alspac_missing_all, item == "Opt_in" & value == "Missing"), n)`; `r pull(filter(alspac_missing_all, item == "Study" & value == "Missing"), n)`; `r pull(filter(alspac_missing_all, item == "Prefer_ecaw" & value == "Missing"), n)`; respectively from top to bottom), responses of _"I don't understand the question"_ (_n_ = `r pull(filter(alspac_missing_all, item == "Less_willing" & value == "I don't understand the question"), n)`; `r pull(filter(alspac_missing_all, item == "Opt_in" & value == "I don't understand the question"), n)`; `r pull(filter(alspac_missing_all, item == "Study" & value == "I don't understand the question"), n)`; `r pull(filter(alspac_missing_all, item == "Prefer_ecaw" & value == "I don't understand the question"), n)`), and responses of _"Unsure"_ (_n_ = `r pull(filter(alspac_missing_all, item == "Less_willing" & value == "Unsure"), n)`; `r pull(filter(alspac_missing_all, item == "Opt_in" & value == "Unsure"), n)`; `r pull(filter(alspac_missing_all, item == "Study" & value == "Unsure"), n)`; `r pull(filter(alspac_missing_all, item == "Prefer_ecaw" & value == "Unsure"), n)`). Agreement with the first question may be slightly inflated due to the format of the questions in this block. Respondents with a highly positive inclination towards ECAWs would be expected to disagree with the first question, but agree with the next three questions. Four respondents agreed with all four statements, suggesting they may have glazed over the word “less” in the first question \protect\footnotemark[2]. Interpreting responses to the second and third question come with a degree of ambiguity as the survey did not specify what was meant by the term “study” \protect\footnotemark[3].
```{=latex}
\end{singlespace}
}
```
